# Regression and model validation
# Heli Helskyaho

*Describe the work this week and summarize your learning.*

- Describe your work and results clearly. 
- Assume the reader has an introductory course level understanding of writing and reading R code as well as statistical methods.
- Assume the reader has no previous knowledge of your data or the more advanced methods you are using.

```{r}
date()
ds <- read.csv("C:/Users/Heli/Heli/HY/Introduction to Open Data Science/Projects/IODS-project/data\\learning2014.csv", header=TRUE)
ds$Points
dim(ds)
str(ds)
summary(ds)

```
The dataset contains 166 rows and  7 columns.
It includes the gender (F/M), the age, the points and some combination variables built using mean. 
These variables and the original variables they are combined are:

# attitude: Aa, Ab, Ac, Ad, Ae, Af
# deep: D03+D11+D19+D27, D07+D14+D22+D30, D06+D15+D23+D31
# surf: SU02+SU10+SU18+SU26, SU05+SU13+SU21+SU29, SU08+SU16+SU24+SU32
# stra: ST01+ST09+ST17+ST25, ST04+ST12+ST20+ST28

Gender is of type chr, age and points are of type int and the rest of the variables are of type num as shown here:
'data.frame':	166 obs. of  7 variables:
 $ X.gender.  : chr  "\"F\"" "\"M\"" "\"F\"" "\"M\"" ...
 $ X.Age.     : int  53 55 49 53 49 38 50 37 37 42 ...
 $ X.attitude.: num  1.5 1.67 1.5 2.17 1.83 ...
 $ X.deep.    : num  3.58 2.92 3.5 3.5 3.67 ...
 $ X.stra.    : num  3.38 2.75 3.62 3.12 3.62 ...
 $ X.surf.    : num  2.58 3.17 2.25 2.25 2.83 ...
 $ X.Points.  : int  25 12 24 10 22 21 21 31 24 26 ...
 
 The minimum age in the dataset is 17, maximum 55. Values for attitude are between 1.000-4.667, for deep between 1.583-4.917, stra 1.250-5.000, and surf 1.583-4.333. The minimum points are 7.00 and the maximum points are 33.00. The table below shows also the 1st quadrant, median, mean, and 3. quadrant for each variable.

   X.gender.             X.Age.       X.attitude.       X.deep.         X.stra.         X.surf.     
 Length:166         Min.   :17.00   Min.   :1.000   Min.   :1.583   Min.   :1.250   Min.   :1.583  
 Class :character   1st Qu.:21.00   1st Qu.:1.500   1st Qu.:3.333   1st Qu.:2.625   1st Qu.:2.417  
 Mode  :character   Median :22.00   Median :1.667   Median :3.667   Median :3.188   Median :2.833  
                    Mean   :25.51   Mean   :1.883   Mean   :3.680   Mean   :3.121   Mean   :2.787  
                    3rd Qu.:27.00   3rd Qu.:2.000   3rd Qu.:4.083   3rd Qu.:3.625   3rd Qu.:3.167  
                    Max.   :55.00   Max.   :4.667   Max.   :4.917   Max.   :5.000   Max.   :4.333  
   X.Points.    
 Min.   : 7.00  
 1st Qu.:19.00  
 Median :23.00  
 Mean   :22.72  
 3rd Qu.:27.75  
 Max.   :33.00  

```{r}
pairs(ds[-1])
```
The scatter plot above describes the relationships between the variables. We have removed gender from the scatter plot.

```{r}
library(GGally)
library(ggplot2)
p <- ggpairs(ds, mapping = aes(), lower = list(combo = wrap("facethist", bins = 20)))
p
```
Above you can see a more advanced plot describing for instance the correlation of different variables to each others and the distribuiton of each variable.
```{r}
# a scatter plot of points versus attitude
library(ggplot2)
# colnames(learning2014)[7] <- "points"

qplot(attitude, Points, data = ds) + geom_smooth(method = "lm")

```

```{r}
my_model <- lm(Points  ~ attitude + deep + Age, data = ds )
summary(my_model)
```
